---
title: "Homework 5"
author: "[Lekh Shetty]{style='background-color: yellow;'}"
toc: true
title-block-banner: true
title-block-style: default
execute: 
  freeze: true
  cache: true
format:
  #html: # comment this line to get pdf
  pdf: 
    fig-width: 7
    fig-height: 7
editor: 
  markdown: 
    wrap: 72
---

[Link to the Github repository](https://github.com/psu-stat380/hw-5)

------------------------------------------------------------------------

::: {.callout-important style="font-size: 0.8em;"}
## Due: Wed, Apr 19, 2023 \@ 11:59pm

Please read the instructions carefully before submitting your
assignment.

1.  This assignment requires you to only upload a `PDF` file on Canvas
2.  Don't collapse any code cells before submitting.
3.  Remember to make sure all your code output is rendered properly
    before uploading your submission.

⚠️ Please add your name to the author information in the frontmatter
before submitting your assignment ⚠️
:::

In this assignment, we will explore decision trees, support vector
machines and neural networks for classification and regression. The
assignment is designed to test your ability to fit and analyze these
models with different configurations and compare their performance.

We will need the following packages:

```{R, message=FALSE, warning=FALSE, results='hide'}
packages <- c(
  "dplyr", 
  "readr", 
  "tidyr", 
  "purrr", 
  "broom",
  "magrittr",
  "corrplot",
  "caret",
  "rpart",
  "rpart.plot",
  "e1071",
  "torch", 
  "luz"
)

#renv::install(packages)
sapply(packages, require, character.only=T)
```

## <br><br><br><br>

## Question 1

::: callout-tip
## 60 points

Prediction of Median House prices
:::

###### 1.1 (2.5 points)

The `data` folder contains the `housing.csv` dataset which contains
housing prices in California from the 1990 California census. The
objective is to predict the median house price for California districts
based on various features.

Read the data file as a tibble in R. Preprocess the data such that:

1.  the variables are of the right data type, e.g., categorical
    variables are encoded as factors
2.  all column names to lower case for consistency
3.  Any observations with missing values are dropped

```{r}
path <- "data/housing.csv"

df <- read_csv(path) %>%
  mutate_if(\(x) is.character(x), as.factor) %>%
  (\(x) {
        names(x) <- tolower(names(x))
        x
    })

df <- na.omit(df)
head(df)
```

------------------------------------------------------------------------

###### 1.2 (2.5 points)

Visualize the correlation matrix of all numeric columns in `df` using
`corrplot()`

```{r}
df %>% 
  select_if(is.numeric) %>% 
  cor() %>% 
  corrplot(type = "upper", method = "color")
```

------------------------------------------------------------------------

###### 1.3 (5 points)

Split the data `df` into `df_train` and `df_split` using `test_ind` in
the code below:

```{r}
set.seed(42)
test_ind <- sample(
  1:nrow(df), 
  floor( nrow(df)/10 ),
  replace=FALSE
)

df_train <- df[-test_ind, ]
df_test  <- df[test_ind, ]

nrow(df_train) + nrow(df_test) == nrow(df)
```

------------------------------------------------------------------------

###### 1.4 (5 points)

Fit a linear regression model to predict the `median_house_value` :

-   `latitude`
-   `longitude`
-   `housing_median_age`
-   `total_rooms`
-   `total_bedrooms`
-   `population`
-   `median_income`
-   `ocean_proximity`

Interpret the coefficients and summarize your results.

```{r}
lm_fit <- lm(median_house_value ~. -households, data = df_train)

summary(lm_fit)
```

------------------------------------------------------------------------

###### 1.5 (5 points)

Complete the `rmse` function for computing the Root Mean-Squared Error
between the true `y` and the predicted `yhat`, and use it to compute the
RMSE for the regression model on `df_test`

```{r}
rmse <- function(y, yhat) {
  sqrt(mean((y - yhat)^2))
}

lm_predictions <- predict(lm_fit, newdata = df_test)

lm_rmse <- rmse(df_test$median_house_value, lm_predictions)
lm_rmse
```

###### 1.6 (5 points)

Fit a decision tree model to predict the `median_house_value` using the
same predictors as in 1.4. Use the `rpart()` function.

```{r}
rpart_fit <- rpart(median_house_value ~. -households, 
                    data = df_train)

rpart_predictions <- predict(rpart_fit, newdata = df_test)

```

Visualize the decision tree using the `rpart.plot()` function.

```{r}
rpart.plot(rpart_fit)
```

Report the root mean squared error on the test set.

```{r}
rpart_predictions <- predict(rpart_fit, newdata = df_test)

rpart_rmse <- rmse(df_test$median_house_value, rpart_predictions)
rpart_rmse

```

------------------------------------------------------------------------

###### 1.7 (5 points)

Fit a support vector machine model to predict the `median_house_value`
using the same predictors as in 1.4. Use the `svm()` function and use
any kernel of your choice. Report the root mean squared error on the
test set.

```{r}
svm_fit <- svm(median_house_value ~. -households,
                data = df_train,
                kernel = "linear",
                cost = 10,
                scale = TRUE)

svm_predictions <- predict(svm_fit, newdata = df_test)

svm_rmse <- rmse(df_test$median_house_value, svm_predictions)
svm_rmse
```

------------------------------------------------------------------------

###### 1.8 (25 points)

Initialize a neural network model architecture:

```{r}
NNet <- nn_module(
    initialize = function(p, q1, q2, q3){
      self$hidden1 <- nn_linear(p, q1)
      self$hidden2 <- nn_linear(q1, q2)
      self$hidden3 <- nn_linear(q2, q3)
      self$output <- nn_linear(q3, 1)
      self$activation <- nn_relu()
      self$sigmoid <- nn_sigmoid()
    },
    forward = function(x){
      x %>% 
      self$hidden1() %>% self$activation() %>% 
      self$hidden2() %>% self$activation() %>% 
      self$hidden3() %>% self$activation() %>% 
      self$output() %>% self$sigmoid()
    }
)
```

Fit a neural network model to predict the `median_house_value` using the
same predictors as in 1.4. Use the `model.matrix` function to create the
covariate matrix and `luz` package for fitting the network with
$32, 16, 8$ nodes in each of the three hidden layers.

```{r}
M <- model.matrix(median_house_value ~ . -households, data = df_train)
```

```{r}
nnet_fit <- NNet %>% 
  setup(
    loss = nn_bce_loss(),
    optimizer = optim_adam,
    #metrics = list(
     #       luz_metric_accuracy()
  ) %>%
  set_hparams(
    p = ncol(M), q1=32, q2=16, q3=8
  ) %>%
  set_opt_hparams(
    lr = 0.1
  ) %>%
  fit(
    data = list(
      model.matrix(median_house_value ~. -households, data = df_train),
      (df_train$median_house_value %>% as.numeric()-1) %>% as.matrix
    ),
    valid_data = list(
      model.matrix(median_house_value ~. -households, data = df_test),
      (df_test$median_house_value %>% as.matrix()-1) %>% as.matrix
    ),
    epochs=50,
    dataloader_options = list(batch_size = 128, shuffle=TRUE),
    verbose = FALSE # Change to TRUE while tuning. But, set to FALSE before submitting
  )

```
Plot the results of the training and validation loss and accuracy.

```{r}
plot(nnet_fit)# Insert your code here
```

Report the root mean squared error on the test set.

```{r}
nnet_predictions <- predict(
    nnet_fit, 
    model.matrix(median_house_value ~ . -households, data = df_test)
)

rmse_nnet <- rmse(df_test$median_house_value, nnet_predictions)
rmse_nnet

```

::: callout-warning
Remember to use the `as_array()` function to convert the predictions to
a vector of numbers before computing the RMSE with `rmse()`
:::

------------------------------------------------------------------------

###### 1.9 (5 points)

Summarize your results in a table comparing the RMSE for the different
models. Which model performed best? Why do you think that is?

``` r
... # Insert your code here
```

<br><br><br><br> <br><br><br><br> ---

## Question 2

::: callout-tip
## 50 points

Spam email classification
:::

The `data` folder contains the `spam.csv` dataset. This dataset contains
features extracted from a collection of spam and non-spam emails. The
objective is to classify the emails as spam or non-spam.

------------------------------------------------------------------------

###### 2.1 (2.5 points)

Read the data file as a tibble in R. Preprocess the data such that:

1.  the variables are of the right data type, e.g., categorical
    variables are encoded as factors
2.  all column names to lower case for consistency
3.  Any observations with missing values are dropped

```{r}
spam <- "data/spambase.csv"
df <- read_csv(spam) %>%
  mutate_if(\(x) is.character(x), as.factor) %>%
  (\(x) {
        names(x) <- tolower(names(x))
        x
    })

df <- na.omit(df)
head(df)
```

------------------------------------------------------------------------

###### 2.2 (2.5 points)

Split the data `df` into `df_train` and `df_split` using `test_ind` in
the code below:

```{r}
set.seed(42)
test_ind <- sample(
  1:nrow(df), 
  floor( nrow(df)/10 ),
  replace=FALSE
)

df_train <- df[-test_ind, ]
df_test  <- df[test_ind, ]
```

Complete the `overview` function which returns a data frame with the
following columns: `accuracy`, `error`, `false positive rate`,
`true positive rate`, between the true `true_class` and the predicted
`pred_class` for any classification model.

```{r}
overview <- function(pred_class, true_class) {
  accuracy <- sum(pred_class == true_class) / length(true_class) # Insert your code here
  error <- 1 - accuracy # Insert your code here
  true_positives <- sum(pred_class=="spam" & true_class=="spam") # Insert your code here
  true_negatives <- sum(pred_class == "nonspam" & true_class=="nonspam") # Insert your code here
  false_positives <- sum(pred_class="spam" & true_class=="nonspam") # Insert your code here
  false_negatives <- sum(pred_class=="nonspam" & true_class=="spam") # Insert your code here
  true_positive_rate <- true_positives / sum(true_class == "spam") # Insert your code here
  false_positive_rate <- false_positives / sum(true_class == "nonspam")  # Insert your code here
  return(
    data.frame(
      accuracy = accuracy,
      error = error,
      true_positive_rate = true_positive_rate,
      false_positive_rate = false_positive_rate
    )
  )
}
```

------------------------------------------------------------------------

###### 2.3 (5 points)

Fit a logistic regression model to predict the `spam` variable using the
remaining predictors. Report the prediction accuracy on the test set.

```{r}
glm_fit <- glm(spam ~ ., data = df_train, family = binomial)
glm_classes <- predict(glm_fit, df_test, type = "response") > 0.5

glm_accuracy <- mean(glm_classes == df_test$spam)
glm_accuracy

```

------------------------------------------------------------------------

###### 2.4 (5 points)

Fit a decision tree model to predict the `spam` variable using the
remaining predictors. Use the `rpart()` function and set the `method`
argument to `"class"`.

```{r}
rpart_fit <- rpart(
  spam ~ ., 
  data = df_train, 
  method = "class"
)

rpart_classes <- predict(rpart_fit, df_test, type = "class")
```

Visualize the decision tree using the `rpart.plot()` function.

```{r}
rpart.plot(rpart_fit)
```

Report the prediction accuracy on the test set.

```{r}
rpart_test_pred <- predict(rpart_fit, df_test, type="class")
rpart_classes <- mean(rpart_test_pred == df_test$spam)
rpart_classes

```

------------------------------------------------------------------------

###### 2.5 (5 points)

Fit a support vector machine model to predict the `spam` variable using
the remaining predictors. Use the `svm()` function and use any kernel of
your choice. Remember to set the `type` argument to `"C-classification"`
**if you haven't** already converted `spam` to be of type `factor`.

```{r}
svm_fit <- svm(spam ~ ., data = df_train, kernel = "linear", type = "C-classification")
```

Report the prediction accuracy on the test set.

```{r}
svm_classes <- predict(svm_fit, newdata = df_test)

accuracy <- mean(svm_classes == df_test$spam)
accuracy

```

------------------------------------------------------------------------

###### 2.6 (25 points)

Using the same neural network architecture as in 1.9, fit a neural
network model to predict the `spam` variable using the remaining
predictors.

::: callout-warning
## Classification vs. Regression

Note that the neural network in **Q 1.9** was a regression model. You
will need to modify the neural network architecture to be a
classification model by changing the output layer to have a single node
with a sigmoid activation function.
:::

Use the `model.matrix` function to create the covariate matrix and `luz`
package for fitting the network with $32, 16, 8$ nodes in each of the
three hidden layers.

```{r}
M <- model.matrix(spam ~ 0+.,data=df_train)
```

```{r}
nnet_fit <- NNet %>% 
  setup(
    loss = nn_bce_loss(),
    optimizer = optim_adam,
  ) %>%
  set_hparams(
    p = ncol(M), q1=32, q2=16, q3=8
  ) %>%
  set_opt_hparams(
    lr = 0.1
  ) %>%
  fit(
    data = list(
      model.matrix(spam ~ 0 + ., data=df_train),
      (df_train$spam %>% as.numeric()-1) %>% as.matrix
    ),
    valid_data = list(
      model.matrix(spam ~ 0 +., data=df_test),
      (df_test$spam %>% as.matrix()-1) %>% as.matrix
    ),
    epochs=50,
    dataloader_options = list(batch_size = 128, shuffle=TRUE),
    verbose = FALSE # Change to TRUE while tuning. But, set to FALSE before submitting

  )
```

------------------------------------------------------------------------

###### 2.7 (5 points)

Summarize your results in a table comparing the accuracy metrics for the
different models.

``` r
... # Insert your code here
```

If you were to choose a model to classify spam emails, which model would
you choose? Think about the context of the problem and the cost of false
positives and false negatives.

<br><br><br><br> <br><br><br><br> ---

## Question 3

::: callout-tip
## 60 points

Three spirals classification
:::

To better illustrate the power of depth in neural networks, we will use
a toy dataset called the "Three Spirals" data. This dataset consists of
two intertwined spirals, making it challenging for shallow models to
classify the data accurately.

::: callout-warning
## This is a multi-class classification problem
:::

The dataset can be generated using the provided R code below:

```{R}
generate_three_spirals <- function(){
  set.seed(42)
  n <- 500
  noise <- 0.2
  t <- (1:n) / n * 2 * pi
  x1 <- c(
      t * (sin(t) + rnorm(n, 0, noise)),
      t * (sin(t + 2 * pi/3) + rnorm(n, 0, noise)),
      t * (sin(t + 4 * pi/3) + rnorm(n, 0, noise))
    )
  x2 <- c(
      t * (cos(t) + rnorm(n, 0, noise)),
      t * (cos(t + 2 * pi/3) + rnorm(n, 0, noise)),
      t * (cos(t + 4 * pi/3) + rnorm(n, 0, noise))
    )
  y <- as.factor(
    c(
      rep(0, n), 
      rep(1, n), 
      rep(2, n)
    )
  )
  return(tibble(x1=x1, x2=x2, y=y))
}
```

------------------------------------------------------------------------

###### 3.1 (5 points)

Generate the three spirals dataset using the code above. Plot $x_1$ vs
$x_2$ and use the `y` variable to color the points.

```{R}
df <- generate_three_spirals()

plot(
  df$x1, df$x2,
  col = df$y,
  pch = 20
)
```

Define a grid of $100$ points from $-10$ to $10$ in both $x_1$ and $x_2$
using the `expand.grid()`. Save it as a tibble called `df_test`.

```{r}
grid <- expand.grid(
  x1 = seq(-10, 10, length.out = 100),
  x2 = seq(-10, 10, length.out = 100)
)

df_test <- as_tibble(grid)

```

------------------------------------------------------------------------

###### 3.2 (10 points)

Fit a classification tree model to predict the `y` variable using the
`x1` and `x2` predictors, and plot the decision boundary.

```{r}
rpart_fit <- rpart(
  y ~ x1 + x2,
  data = df,
  method = "class"
)

rpart_classes <- predict(rpart_fit, df_test, type = "class")

rpart.plot(rpart_fit)
```

Plot the decision boundary using the following function:

```{r}
plot_decision_boundary <- function(predictions){
  plot(
    df_test$x1, df_test$x2, 
    col = predictions,
    pch = 0
  )
  points(
    df$x1, df$x2,
    col = df$y,
    pch = 20
  )
}
```

```{r}
plot_decision_boundary(rpart_classes)
```

------------------------------------------------------------------------

###### 3.3 (10 points)

Fit a support vector machine model to predict the `y` variable using the
`x1` and `x2` predictors. Use the `svm()` function and use any kernel of
your choice. Remember to set the `type` argument to `"C-classification"`
**if you haven't** converted `y` to be of type `factor`.

``` r
svm_fit <- ... # Insert your code here
svm_classes <- ... # Insert your code here
plot_decision_boundary(svm_classes)
```

```{r}
svm_fit <- svm(y ~ ., data = df, kernel = "radial", type = "C-classification")
svm_classes <- predict(svm_fit, df_test)
plot_decision_boundary(svm_classes)

```

------------------------------------------------------------------------

::: callout-warning
## Instructions

For the next questions, you will need to fit a series of neural
networks. In all cases, you can:

-   set the number of units in each hidden layer to 10
-   set the output dimension `o` to 3 (remember this is multinomial
    classification)
-   use the appropriate loss function for the problem (**not
    `nn_bce_loss`**)
-   set the number of epochs to $50$
-   fit the model using the `luz` package

You can use any optimizer of your choice, but you **will need to tune
the learning rate for each problem**.
:::

###### 3.4 (10 points)

Fit a neural network with **1 hidden layer** to predict the `y` variable
using the `x1` and `x2` predictors.

```{r}
NN1 <- nn_module(
  initialize = function(p, q1, o){
    self$hidden1 <- nn_linear(p, q1) 
    self$output <- nn_linear(q1, o) 
    self$activation <- nn_sigmoid() 
  },
  forward = function(x){
    x %>% 
      self$hidden1() %>% 
      self$activation() %>% 
      self$output()
  }
)

fit_1 <- NN1 %>% 
  setup(
    loss = nn_cross_entropy_loss(),
    optimizer = optim_adam
  ) %>%
  set_hparams(
    p = ncol(df_test), q1=10, o=3
  ) %>%
  set_opt_hparams(
    lr = 0.001
  ) %>%
  fit(
    data = list(
      df %>% select(x1, x2) %>% as.matrix,
      df$y %>% as.integer
    ),
    epochs=50,
    dataloader_options = list(batch_size = 200, shuffle = TRUE),
    verbose = FALSE
  )
```

In order to generate the class predictions, you will need to use the
`predict()` function as follows

```r
test_matrix <- df_test %>% select(x1, x2) %>% as.matrix

fit_1_predictions <- predict(fit_1, test_matrix) %>% 
  argmax(2) %>% 
  as.integer()
```

Plot the results using the `plot_decision_boundary()` function.

------------------------------------------------------------------------

###### 3.5 (10 points)

Fit a neural network with **0 hidden layers** to predict the `y`
variable using the `x1` and `x2` predictors.

```r
NN0 <- nn_module(
  initialize = function(p, o){
    self$output <- nn_dense(p, o) # Insert your code here
  },
  forward = function(x){
    x %>% 
    self$output() # Insert your code here
  }
)

fit_0 <- NN0 %>% 
  setup(...) %>%
  set_hparams(...) %>%
  set_opt_params(...) %>%
  fit(...)
```

Plot the results using the `plot_decision_boundary()` function.

------------------------------------------------------------------------

###### 3.6 (10 points)

Fit a neural network with **3 hidden layers** to predict the `y`
variable using the `x1` and `x2` predictors.

``` r
NN2 <- nn_module(
  initialize = function(p, q1, q2, o){
    ... # Insert your code here
  },
  forward = function(x){
    x %>% 
    ... # Insert your code here
  }
)

fit_2 <- NN3 %>% 
  setup(...) %>%
  set_hparams(...) %>%
  set_opt_params(...) %>%
  fit(...)
```

Plot the results using the `plot_decision_boundary()` function.

------------------------------------------------------------------------

###### 3.7 (5 points)

What are the differences between the models? How do the decision
boundaries change as the number of hidden layers increases?

------------------------------------------------------------------------

::: {.hidden unless-format="pdf"}
\pagebreak
:::

<br><br><br><br> <br><br><br><br> ---

::: {.callout-note collapse="true"}
## Session Information

Print your `R` session information using the following command

```{R}
sessionInfo()
```
:::
